Propositions as Sessions, Mechanically

### 2015-02-24 paper/paper.tex
Structuring the discussion of the development.

This project describes work towards the development of a framework for
formalising lambda calculus type systems with session types using interactive
theorem proving (in particular, the system Coq). The framework should enable a
whole class of related languages and type systems to be formalised with little
extra proof effort. Inspired by previous work on a mechanised framework for
$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
the lambda calculus to provide a basis on which session types for mainstream
programming languages can be studied, whilst taking advantage of previous
proof effort and specialising only for those parts unique to the particular
system. The framework is heavily based on calculi developed by
\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
classical linear logic (CP) and a functional language with session types (GV).

@@

The definition of session types will be important for the remainder:
\input{../proposal/session-types.tex}

@@

They represent dual intuitionistic linear logic as a sequent calculus
providing logical connectives which have session-based interpretations in
$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL, which have
only two alternatives (cf. multiple alternatives in the standard definition in
\S~\ref{sec:pre}):
\input{../proposal/sill-branch-choice-rules}

Note that branch and choice both have some notion of branching (R rules for
$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
a session-based setting, since the intutition for \& is that of
\emph{offering} (receiving) a choice between alternatives rather than
\emph{making} (sending) the choice. Here, however, the intuition should be
extended to account for its location in the typing judgement. The channel is
on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
consider the context as services being provided to the typed process. Thus,
the type of channels in the context is really the type of the other endpoint
of the channel, and the typed process behaves according to the dual
protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
represent the dual of a protocol $A$ as defined by
\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
alternatives (in this case, $\inrterm$) and then behaving according to the
selected protocol (in this case, protocol $B^\bot$).

@@

I describe the formalisation as it currently stands in this section.

@@

There remains much to be done.

>>>>

This project describes work towards the development of a framework for
formalising lambda calculus type systems with session types using interactive
theorem proving (in particular, the system Coq). The framework should enable a
whole class of related languages and type systems to be formalised with little
extra proof effort. Inspired by previous work on a mechanised framework for
$\pi$-calculus type systems~\cite{Gay:2001:FFP}, the current work focuses on
the lambda calculus to provide a basis on which session types for mainstream
programming languages can be studied, whilst taking advantage of previous
proof effort and specialising only for those parts unique to the particular
system. The starting point for the framework is the calculi developed by
\citeauthor{Wadler:2014}~\cite{Wadler:2014} describing a relationship between
classical linear logic (CP) and a functional language with session types (GV).

@@

The definition of session types will be important for the remainder:
\input{../paper/session-types.tex}

@@

\input{../paper/sill-branch-choice-rules}

They represent dual intuitionistic linear logic as a sequent calculus
providing logical connectives which have session-based interpretations in
$\pi$DILL. Consider their rules for branch and choice in $\pi$DILL in
Figure~\ref{fig:sill-branching}, which have only two alternatives
(cf. multiple alternatives in the standard definition in \S~\ref{sec:pre}).

Note that branch and choice both have some notion of branching (R rules for
$\oplus$/L rules for $\with$) and at least for \& this is counter-intuitive in
a session-based setting, since the intuition for \& is that of
\emph{offering} (receiving) a choice between alternatives rather than
\emph{making} (sending) the choice. Here, however, the intuition should be
extended to account for its location in the typing judgement. The channel is
on the left of the sequent, and \citeauthor{Caires:2010:STI} suggest one
consider the context as services being provided to the typed process. Thus,
the type of channels in the context is really the type of the other endpoint
of the channel, and the typed process behaves according to the dual
protocol. For example, considering $\SILLbranchLTwo$, and taking $A^\bot$ to
represent the dual of a protocol $A$ as defined by
\citeauthor{Wadler:2014}~\cite{Wadler:2014}, the typed process is behaving
according to the protocol $\BChoice{A^\bot}{B^\bot}$, selecting from
alternatives (in this case, $\inrterm$) and then behaving according to the
selected protocol (in this case, protocol $B^\bot$).

@@

The formalisation was performed using the Coq proof
assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
challenge~\cite{Aydemir:2005:MMM}, the development utilises available
libraries and infrastructure as much as possible. In particular,
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
library for handling, among other things, association lists (environments) and
free variable representation. The current development uses the version of the
library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
has been updated for Coq version 8.4pl4. One goal of the formalisation effort
was to minimise the extent to which the development differs from the
pen-and-paper presentation and avoid questions of adequacy of the
representation. Some changes were made to ease the mechanisation and are
highlighted where they occur.

The locally nameless first-order representation for binders and free variables
was adopted for both GV and CP, following
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
variant prevents variable-capture issues when compared with other first-order
approaches. Alternatively, I could have chosen a higher-order abstract syntax
(HOAS) approach such as the parametric version (PHOAS) described
by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
on the metalanguage (Coq, in this case) for handling binders and substitution
so has less overhead in managing syntax than the locally nameless
approach. However, it was less clear how to handle linearity using this
approach and has the disadvantage of being able to represent terms not valid
in the language being mechanised (so called ``exotic'' terms). Further, a
higher-order approach would prevent the use of the Metatheory library which
provides a number of useful tactics for manipulating typing contexts. Previous
efforts to mechanise \fpop~\cite{Park:2014:MMW} illustrate how to handle a
language containing both linear and non-linear contexts. However, to stay true
to CP/GV, I maintained a single context. Interestingly, it became clear that
the mechanised version of CP required explicit weakening and contraction to
facilitate the reduction rules.

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\cite{Aydemir:2008:EFM}, which excludes a finite set of
variables from being considered as the binder. In contrast, the traditional
``exists-fresh'' approach, where the binder is only required to be fresh
within the abstraction's body, does not produce strong enough induction
principles in some cases. This change will necessitate changes to the typing
rules of GV and CP. Consider for instance the Connect rule in
GV~\cite{Wadler:2014} (C) and its corresponding cofinite quantification
version (CCO): \input{with-connect-rule}

The conclusion specifies a variable in (C) but a type in (CCO) (de Bruijn
index rather than a name). Thus, in the premise for (CCO) one can choose a
suitably fresh $x$ in each premise. Note that, $L$ can be chosen such that
$\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$, where $\fv(M)$ are the
set of free variables in term $M$, and $x$ can be the same in both
premises. Similar changes are made to the rules for CP (see
\S~\ref{sec:cp}). These changes do not alter the typing of processes nor the
semantics of the translation from GV to CP. An equivalence between
``exists-fresh'' and cofinite quantification definitions of the simply-typed
$\lambda$-calculus is given by
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
result could be derived straightforwardly (but tediously) for the CP/GV
system.

\subsection{The process calculus CP}\label{sec:cp}

\begin{comment}
FIGURES

propositions

processes

typing judgements

equivalences and axiomatic cut

principal cut rules

commuting conversions
\end{comment}

Formalising the process calculus CP provided many challenges for the
metatheory library since to my knowledge, the library has never been applied
to a process calculus before. So for instance, in the lambda calculus one
typically does not work up to permutation of binders, whereas in a process
calculus these kinds of structural equivalences are standard.

Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
polymorphism the reduction rules are unnecessary for the translation. Future
work is to support reduction under these terms.

\subsection{The functional language GV}\label{sec:gv}

\begin{comment}
FIGURES

types

terms

typing judgements
\end{comment}

\subsection{CPS translation GV to CP}

Currently the formalisation does not handle this. Future work.

for an unbound index so this decision is sound. A consequence is that I must
define a notion of opening language terms involving binders as described
by~\citeauthor{Aydemir:2008:EFM}. For example, an abstraction with body term
$t$ which is well-typed if $t$ is well-typed when its unbound index,
representing the argument to the abstraction, is replaced by some fresh
variable $x$, written $t^x$. Some terms in GV require multiple binders but I
believe it will be straightforward to extend this notion. Unfortunately, I
must define similar operations on CP terms and handle substitution and opening
on them as well as propositions which can accept propositional variables. I
hope to be able to generalise as much as possible with the help of the
Metatheory library. say something about gmeta here and its automatic
generation of these infrastructure lemmas

While the aim is to be faithful to \citeauthor{Wadler:2014}'s description, I
may need to alter the approach slightly to avoid tricky formulations. For
instance, instead of providing $n$-ary branch and choice in GV, I provide
only binary versions of these operators to simplify the development. This
change is not restrictive however, since CP itself provides only binary
versions of plus ($\oplus$) and with ($\with$) constructs. Likewise, on a more
technical note, I amalgamate all types into one inductive definition and
provide a predicate for restricting typing rules to consider only valid
session types. \citeauthor{Wadler:2014} defines GV types as a set of mutually
recursive definitions; session types are types which may contain types as
subcomponents (arguments to send, for example). Unfortunately, handling
mutually inductive definitions in Coq can be quite involved; requiring one to
either rely on the Coq system to provide a stronger mutual induction principle
or defining one manually. Such a definition is possible, but it complicates
elimination via the induction principle.

Another aspect of the encoding is how to define terms. The intrinsic encoding
described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
approach, indexing terms by their type so as to prevent ill-formed terms from
being constructed. However, in the encoding for GV there are issues with using
this encoding: $(1)$ it is not immediately clear how to handle linear contexts
using the de Bruijn variable encoding, since in the intrinsic setting the
environment supports weakening; $(2)$ an intrinsic encoding of terms would be
complicated by the need to enforce that session types occur in certain
instances (by use of a predicate); and $(3)$ well-typed terms require extra
assumptions about binder freshness which cannot reasonably be expressed as a
function type. For these reasons, an intrinsic approach does not offer much
benefit since a well-typed term relation would still need to be
defined. Therefore, I chose an extrinsic encoding for terms and define
well-typed terms as a separate inductive type.

\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} note that the size of
their language proof infrastructure is proportional to the number of binding
constructs. In the case of a simply typed lambda calculus this is not onerous,
but GV has four binding constructs, CP has six and propositions have two. It
may be true that this approach results in too many auxiliary definitions and
operations not relevant to the actual properties one wishes to prove. However,
I believe that once this infrastructure has been defined, the system can be
extended without concern for these binding constructs.

\citeauthor{Park:2014:MMW} describe a technique for
removing nonlinear contexts from typing judgements. While their work extended
to \fpop it is not clear how to handle an environment containing both
non-linear and linear types as in GV. I wish to maintain as close a
relationship as possible to the paper system presented by
\citeauthor{Wadler:2014}, so separating out the non-linear and linear
components (as in \fpop) is not an option at this stage.

I describe the formalisation as it currently stands in this section. The use
of the metatheory library from UPenn provided useful tactics and properties
regarding freshness of free variables and environments.

@@

In addition to that already mentioned: there remains much to be done.
